Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-22 documentation/userguide.tex
Add documentation on how to build your own problem.

@@

\chapter{Code Structure}

@@

The \texttt{source} directory contains all of the Fortran routines needed to build a CASTRO 
executable file. It is not normally built directly, but you can type \texttt{make} in that directory
to build a vanilla executable built with the default settings.

@@

The \texttt{job\_scripts} directory contains utilities that are used for building and running
problem setups on various machine architectures.

@@

@@

Each test directory is set up with a uniform pattern. The \texttt{source/} subdirectory includes all of the source code
needed to build the test. The \texttt{compile/} directory contains the \texttt{makefile}, which builds the executable and creates
the inputs files needed for the simulation. The \texttt{run\_test.sh} script executes the test by creating a subdirectory
for each parameter set and then submitting it with a batch submission script (for a vanilla Linux installation,
this will use the \texttt{batch} utility; for a limited number of clusters and supercomputers, 
there will be recommended options and specialized submission scripts). The \texttt{analysis.py} Python script
is used to analyze the output and generature figures. In summary, to run a test, perform the following series of commands:
\begin{itemize}
  \item \texttt{cd compile}
  \item \texttt{make}
  \item \texttt{cd ..}
  \item \texttt{bash run\_test.sh}
  \item \texttt{python analysis.py} (after the execution has completed)
\end{itemize}

>>>>

\usepackage[usenames]{color}

% For inserting code snippets;
% use the lstlisting environment to access.
% emacs doesn't parse $'s in code sections 
% propertly, so that is why I have sprinkled 
% some %$'s through the code. The coloring 
% on the environment itself is still broken though.
\usepackage{listings}

\definecolor{gray}{rgb}       {0.8,0.8,0.8}
\definecolor{light-blue}{rgb} {0.8,0.8,1.0}
\definecolor{light-green}{rgb}{0.8,1.0,0.8}
\definecolor{light-red}{rgb}  {1.0,0.9,0.9}

\lstset{
  basicstyle=\small\ttfamily,
  frame=shadowbox, 
  rulesepcolor=\color{gray},                                                             
  backgroundcolor=\color{white}
}

@@

A note on machine compatibility: this software is currently only developed and tested on
machines that run Linux and come with the common GNU flavors of most pieces of software
like \texttt{sed}. We do not guarantee that this software will work out of the box on OS X
or other Unix derivatives; however, we generally try to avoid GNU-specific options, so that 
trying to run this software on OS X will not be too painful. With that being said, this 
software does heavily rely on \texttt{bash} and at present there are no plans to maintain
compatibility with any other shell.

\chapter{Directory Layout}

@@

The \texttt{source} directory contains all of the Fortran routines needed to build a CASTRO 
executable file. It is not normally built directly, but you can type \texttt{make} in that directory
to build a vanilla executable with the default settings.

@@

The \texttt{job\_scripts} directory contains utilities that are used for running
problem setups on various machine architectures.

@@

\chapter{How to Build a Problem}

This chapter describes the necessary steps you need to take to build your own problem setup,
and describes the underlying code that facilitates the job submission. If you are interested in
doing white dwarf mergers and want to use the default settings, read Section \ref{sec:standard_problems}.
If you are interested in setting up a different problem but using this software's job submission 
system, you can do so; read Section \ref{sec:non-standard_problems} for more information.

\section{Standard Problems}
\label{sec:standard_problems}

At a minimum, your working directory needs a \texttt{compile/} subdirectory with a \texttt{makefile}
in it, and some \texttt{bash} script that you run to submit the jobs. If you want to build 
with the default settings, then your makefile needs only to have the line:
\begin{lstlisting}
  include $(WDMERGER_HOME)/source/GNUmakefile
\end{lstlisting}%$
This includes the main makefile from the \texttt{source/} directory in \texttt{wdmerger}. 
The main makefile sets all of our build options (compiler, number of space dimensions, 
equation of state, etc.) and then includes the CASTRO makefile, which then runs as normal.
If you want to adjust the build options, then simply add a line before this include. As an 
example, if you want to build using a gamma-law equation of state, set:
\begin{lstlisting}
  EOS_dir = gamma_law_general/

  include $(WDMERGER_HOME)/source/GNUmakefile 
\end{lstlisting}%$

The \texttt{make} variables in the main makefile are all set conditionally, so as long 
as you set your variable before including the main makefile, it will be built as you expect.

Once your makefile is setup, you can simply type \texttt{make} (or \texttt{make -j} 
to use more processors and speed up the operation) at the command line. 
If you're using Cray compilers on a supercomputer this will take several minutes;
with GNU compilers it should be quick. When everything is compiled, if it was
successful you should have an executable that looks like \texttt{Castro3d.Linux.Cray.Cray.MPI.ex}
in the \texttt{compile} directory.

Now \texttt{cd} up one level, and use your favorite text editor to create a simple 
\texttt{bash} script, which for example purposes we will call \texttt{run.sh}.
The first thing we should do is tell the script where to look to pull in 
all the helper functions we'll need to execute a job:

\begin{lstlisting}
source $WDMERGER_HOME/job_scripts/run_utils.sh
\end{lstlisting}%$

Decide what directory you want to build the problem in. It is expected 
to be a subdirectory of the \texttt{results/} directory, which will be
automatically created for you:
\begin{lstlisting}
dir=results/test_problem/
\end{lstlisting}
Now, optionally decide how many processors you want to run with, and 
how long you want the job to run:
\begin{lstlisting}
nprocs=128
walltime=4:00:00
\end{lstlisting}
If you do not, the job submission script will default to running the job
using one node for one hour. Finally, run your job:
\begin{lstlisting}
run $dir $nprocs $walltime
\end{lstlisting}%$
That is all you need to submit a job. Return to the terminal, and type \texttt{bash run.sh}
and your job will execute. The helper function \texttt{run} does several things. First, 
it creates the directory named \texttt{\$dir}. Then, it copies into \texttt{\$dir} the \texttt{inputs}
and \texttt{probin} files from the root \texttt{wdmerger} source directory; these are the runtime
files CASTRO needs to determine what your problem setup will be. They contain the default 
parameters for merger runs. It also copies any auxiliary files one might need, like the 
table for the equation of state, as well as the actual CASTRO executable you just built in the 
\texttt{compile/} directory. Finally, it builds a script that you'll use for actually submitting 
the job to the batch submission system on the machine you're on, and then submits that job. 
Thus for a typical setup, your run directory will contain the following files when you submit:
\begin{itemize}
  \item \texttt{Castro3d*.ex}
  \item \texttt{helm\_table.dat}
  \item \texttt{inputs}
  \item \texttt{probin}
  \item \texttt{run\_script}
\end{itemize}
Typically you'll want to run multiple jobs at once, that vary things like the number of processors 
or the runtime parameters. For example, suppose we want to run the same problem but we want to 
explore what happens when we change the rotational period of the system. The main helper functions 
will automatically handle this for you as long as your variable is part of a valid namespace in the inputs files.
The orbital period is handled by the inputs parameter \texttt{castro.rotational\_period}. Therefore 
we simply need to update \texttt{rotational\_period}. A sample script follows:

\begin{lstlisting}
source $WDMERGER_HOME/job_scripts/run_utils.sh

nprocs=512
walltime=12:00:00

for period in 25.0 50.0 75.0 100.0
do
  dir=results/period$period/
  rotational_period=$period
  stop_time=$period
  run $dir $nprocs $walltime
done
\end{lstlisting}%$
This will create the subdirectories \texttt{results/period25.0} through \texttt{results/period100.0}
and in each directory the inputs file will have the corresponding \texttt{castro.rotational\_period}.
Each job will stop after one rotational period has been completed. When all have been completed, you should 
type \texttt{bash run.sh} once more; this will move all output files (checkpoints, plot files, diagnostics)
to the \texttt{output/} subdirectory of that problem location, and on supported HPC systems will also copy these 
files to the long-term storage system (see Section \ref{sec:archiving} for more details). 
Generally our analysis scripts expect all the output to be in the 
\texttt{output/} subdirectory. Also, if the machine crashes before your run completes, running \texttt{bash run.sh}
again will automatically restart your job from the most recent checkpoint.

\section{Non-Standard Problems}
\label{sec:non-standard_problems}

If you want to build your own problem setup, you can do that too. The makefile and job submission 
requirements are the same as in Section \ref{sec:standard_problems}, but you will need to provide 
your own source files. They should all go in a \texttt{source/} subdirectory in your problem location.
At minimum, you'll need a \texttt{Prob\_3d.f90} file (or the equivalent for lower dimensions) and a 
probin file. The probin file can be empty, it just needs to be there because otherwise the helper 
routines will make a copy of the default probin file. To get a sense of how this is done, look at 
some of the existing test problems, in particular the \texttt{kelvin\_helmholtz} test. To tell the 
code that you want to compile this source code instead of the default code, make a file called 
\texttt{full\_source} and place it in the \texttt{source/} directory (you can just use \texttt{touch}
to create an empty file). Then follow the rest of the instructions as normal, starting with the 
compile step.

The only source code that will be built from the default location is \texttt{sum\_integrated\_quantities.cpp},
which controls the diagnostic output. Our version is a bit more extensive than what comes in CASTRO
by default, so we include this in every build.

@@

Each test directory is set up with a uniform pattern. The \texttt{source/} subdirectory includes all of the source code
needed to build the test. The \texttt{compile/} directory contains the \texttt{GNUmakefile}, which builds the executable.
The \texttt{run\_test.sh} script executes the test by creating a subdirectory
for each parameter set and then submitting it with a batch submission script (for a vanilla Linux installation,
this will use the \texttt{batch} utility; for a limited number of clusters and supercomputers, 
there will be recommended options and specialized submission scripts). The \texttt{analysis.py} Python script
is used to analyze the output and generate figures. In summary, to run a test, perform the following series of commands:
\begin{itemize}
  \item \texttt{cd compile}
  \item \texttt{make}
  \item \texttt{cd ..}
  \item \texttt{bash run\_test.sh}
  \item \texttt{bash run\_test.sh} (after the execution has completed, to archive output)
  \item \texttt{python analysis.py} (after the execution has completed)
\end{itemize}
